Short-term neuronal and synaptic plasticity act in synergy for deviance detection in spiking networks

### 2023-03-20 paper-1/elife.tex
Set to 1.5 line spacing
Added WPI funding ack
Fixed hat t_j appearance
Aligned fig file naming & fix references in tex

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% ELIFE ARTICLE TEMPLATE
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% PREAMBLE 
\documentclass[9pt,lineno]{elife}
% Use the onehalfspacing option for 1.5 line spacing
% Use the doublespacing option for 2.0 line spacing
% Please note that these options may affect formatting.
% Additionally, the use of the \newcommand function should be limited.

\usepackage{lipsum} % Required to insert dummy text
\usepackage[version=4]{mhchem}
\usepackage{siunitx}
\DeclareSIUnit\Molar{M}

@@

We modeled neurons as leaky integrate-and-fire units whose membrane potential $V_m$ at time $t$ followed
\begin{equation}
    \tau_m \frac{dV_m}{dt} = (V_{rest}-V_m) + I_{syn}(t)
\end{equation}
with $\tau_m = 30$~ms the membrane time constant and $V_{rest} = -60$~mV the resting membrane potential. Synaptic currents were modeled in a conductance-based manner, following
\begin{equation}
    I_{syn} = g_e(E_e-V_m) + g_i(E_i-V_m)
\end{equation}
with $E_e = 0$~mV and $E_i = -100$~mV the excitatory and inhibitory reversal potentials, respectively. Synaptic conductances evolved according to
\begin{align} 
    \tau_e \frac{dg_e}{dt} &= -g_e + \sum_{j \in \boldsymbol E} U x_j w \delta(t - \hat{t_j}) \nonumber \\
    \tau_i \frac{dg_i}{dt} &= -g_i + \sum_{j \in \boldsymbol I} w \delta(t - \hat{t_j}) \label{eq:gsyn}
\end{align}
with $\tau_e = 2$~ms the excitatory time constant, echoing AMPA receptor dynamics \citep{Hausser1997-cn}, $\tau_i = 4$~ms the inhibitory time constant, echoing GABA-A receptor dynamics \citep{Destexhe1994-oc}, $\boldsymbol E$ and $\boldsymbol I$ the sets of presynaptic excitatory and inhibitory neurons, respectively, $w$ the synaptic weight, $\delta(\cdot)$ the Dirac delta function, and $\hat{t}$ the spike times.

@@

\begin{figure}
    \includegraphics[width=\linewidth]{fig10}
    \caption{%
        Suppression acts in spatially distinguishable subpopulations.
        \textbf{A} Response size $\R{A}{}$ as a function of distance from the center of the stimulation site, evaluated across all networks and stimuli in 200 bins between 0 mm and 6.5 mm, and normalized by the response size in recovered networks (initial trial). Each of the 200 data points represents the mean response across all relevant trials of all neurons in the corresponding distance bin. Left, spatial illustration, using all data with a fictive stimulus location. Right, same data plotted as lines. The vertical line indicates the chosen local/global divider.
        \textbf{B} As in panel A, but showing $\T{A}{}$ without normalization.
        \textbf{C} As in panel A, but showing $\D{A}{}$ without normalization.
        \textbf{D} Mean $\T{A}{}$ in neurons <2 mm from the stimulation site (local) and more distant neurons (global). Data are averaged across relevant neurons and trials in each network and stimulus, leaving 120 data points to be summarized by box plot. Comparisons of interest are between \std{} and \dev{} in the local subset, and between \msc{} and \dev{} in the global subset.
        \textbf{E} Corresponding mean $\D{A}{}$ in local and global portions of the network.
    }
    \label{fig:10}
\end{figure}

Finally, we sought to confirm whether ``local'' and ``global'' suppression, which we defined so far by the latency of response, did in fact reflect a spatial property of the network. To do so, we assessed the distance between the center of the target stimulation site and each neuron in all networks and stimulus combinations, and calculated $\R{A}{}$, $\T{A}{}$ and $\D{A}{}$ as a function of this distance. \FIG{10}A-c shows that there were indeed clear spatially defined differences between \dev{}, \msc{} and \std{}. Responses, normalized by the response to stimulation in the recovered networks (\FIG{10}A), clearly spread out farthest in \dev{}, and least in \std{}, despite a nearly indistinguishable response at the immediate stimulation site. For thresholds, $\T[std]{A}{}$ (\FIG{10}B) was much stronger near the stimulation site, and somewhat weaker distally, than either of the less frequent paradigms. A similar pattern emerged with $\D{A}{}$ (\FIG{10}C), though here, the local patch covered a larger area, as we might expect from a synaptic effect.

To quantify this statistically, we split the neurons into local and distal (global) subsets, cutting at a radius of 2 mm from the stimulus site, yielding an average local subset size of 226.8 $\pm$ 13.2 neurons, out of 1000. We then assessed the difference of the mean values of $\T{A}{}$ and $\D{A}{}$ across the two subsets in each sequence, as shown in \FIG{10}D. In keeping with the previously stated analysis, we found that local suppression was higher in \std{} than in \dev{} both for $\T{A}{}$ (T = 482, p = 8.31e-17; one-sided Wilcoxon, n = 120) and for $\D{A}{}$ (T = 735, p = 1.71e-14), indicating that the network response to the non-target stimulus in the \dev{} sequence was indeed kept in check by high local suppression. Similarly, we found that global suppression was greater in \msc{} than in \dev{}, again for both $\T{A}{}$ (T = 1.09e+03, p = 1.37e-11) and $\D{A}{}$ (T = 927, p = 7.27e-13), indicating that the \msc{} sequence produced more suppression across the wider network, leading to smaller target responses than in the \dev{} sequence. Incidentally, this greater global suppression, caused by the varied stimulation paradigm, was also reflected in greater local suppression around the target stimulus site for $\T{A}{}$ (T = 831, p = 1.15e-13) and $\D{A}{}$ (T = 1.46e+03, p = 6.13e-09), providing an additional pathway for a comparatively large response in \dev{}.

@@

\begin{figure}
    \includegraphics[width=\linewidth]{fig9}
    \caption{%
        Illustration of the origin of DD in our model.
        \textbf{A} Left: In an oddball context, the regular inputs are encoded in the network by high local suppression, which reduces their overall response, leading to low global suppression. Deviant inputs encounter low local and global suppression and therefore evoke a large response. Right: In an irregular context, the network frequently encounters rare stimuli with low local suppression, leading to large overall responses and high global suppression, with these two (global response and suppression) counter-balancing each other. Deviant inputs encounter this high global suppression and therefore evoke smaller responses.
        \textbf{B} Adding additional suppression (e.g., STD) affects these situations differently. Left: In an oddball context, the additional suppression further reduces the regular response, permitting even greater recovery of the wider network. Right: In an irregular context, global suppression is instead increased, and all responses reduced.
    }
    \label{fig:9}
\end{figure}

@@

In this study, we showed that non-synaptic neuronal adaptation can induce true deviance detection in a randomly wired, but spatially organized network of simple leaky integrate-and-fire neurons. Neuronal adaptation achieves this, as illustrated in Figure 10a, by suppressing the response of neurons local to the frequently stimulated site. This drastically reduces the response to frequent stimulations at an early stage of signal propagation through the network, allowing the more distant parts of the network to recover from adaptation and respond vigorously to infrequent deviant stimuli. By contrast, a more varied paradigm including inputs at multiple stimulation sites allows local suppression to partly recover between stimulus presentations, resulting in a greater response through greater recruitment of the wider network. As a result, the network becomes more adapted, diminishing the response to each individual stimulus. Deviance detection, then, is a side effect of learning and adapting to the regularity of the frequently presented stimulus, which leaves the wider network less adapted and more responsive than usual.

@@

Our second main finding is that synaptic short-term depression, which acts in similar fashion to neuronal adaptation both in terms of how it is induced and in how it affects activity flow in the network, enhances the DD effect of neuronal adaptation, and vice versa. We demonstrated that this synergy is a result of increased local suppression of the frequently presented stimulus by STD, which decreases global suppression by both mechanisms in oddball sequences and enhances the deviant response, as illustrated in Figure 10b. Conversely, without strong local suppression in control sequences, global suppression is increased, leading to smaller individual responses.

@@

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% ELIFE ARTICLE TEMPLATE
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% PREAMBLE 
\documentclass[9pt,lineno,onehalfspacing]{elife}
% Use the onehalfspacing option for 1.5 line spacing
% Use the doublespacing option for 2.0 line spacing
% Please note that these options may affect formatting.
% Additionally, the use of the \newcommand function should be limited.

\usepackage[version=4]{mhchem}
\usepackage{siunitx}
\DeclareSIUnit\Molar{M}

@@

We modeled neurons as leaky integrate-and-fire units whose membrane potential $V_m$ at time $t$ followed
\begin{equation}
    \tau_m \frac{dV_m}{dt} = (V_{rest}-V_m) + I_{syn}(t)
\end{equation}
with $\tau_m = 30$~ms the membrane time constant and $V_{rest} = -60$~mV the resting membrane potential. Synaptic currents were modeled in a conductance-based manner, following
\begin{equation}
    I_{syn} = g_e(E_e-V_m) + g_i(E_i-V_m)
\end{equation}
with $E_e = 0$~mV and $E_i = -100$~mV the excitatory and inhibitory reversal potentials, respectively. Synaptic conductances evolved according to
\begin{align} 
    \tau_e \frac{dg_e}{dt} &= -g_e + \sum_{j \in \boldsymbol E} U x_j w \delta(t - \hat{t}_j) \nonumber \\
    \tau_i \frac{dg_i}{dt} &= -g_i + \sum_{j \in \boldsymbol I} w \delta(t - \hat{t}_j) \label{eq:gsyn}
\end{align}
with $\tau_e = 2$~ms the excitatory time constant, echoing AMPA receptor dynamics \citep{Hausser1997-cn}, $\tau_i = 4$~ms the inhibitory time constant, echoing GABA-A receptor dynamics \citep{Destexhe1994-oc}, $\boldsymbol E$ and $\boldsymbol I$ the sets of presynaptic excitatory and inhibitory neurons, respectively, $w$ the synaptic weight, $\delta(\cdot)$ the Dirac delta function, and $\hat{t}$ the spike times.

@@

\begin{figure}
    \includegraphics[width=\linewidth]{fig9}
    \caption{%
        Suppression acts in spatially distinguishable subpopulations.
        \textbf{A} Response size $\R{A}{}$ as a function of distance from the center of the stimulation site, evaluated across all networks and stimuli in 200 bins between 0 mm and 6.5 mm, and normalized by the response size in recovered networks (initial trial). Each of the 200 data points represents the mean response across all relevant trials of all neurons in the corresponding distance bin. Left, spatial illustration, using all data with a fictive stimulus location. Right, same data plotted as lines. The vertical line indicates the chosen local/global divider.
        \textbf{B} As in panel A, but showing $\T{A}{}$ without normalization.
        \textbf{C} As in panel A, but showing $\D{A}{}$ without normalization.
        \textbf{D} Mean $\T{A}{}$ in neurons <2 mm from the stimulation site (local) and more distant neurons (global). Data are averaged across relevant neurons and trials in each network and stimulus, leaving 120 data points to be summarized by box plot. Comparisons of interest are between \std{} and \dev{} in the local subset, and between \msc{} and \dev{} in the global subset.
        \textbf{E} Corresponding mean $\D{A}{}$ in local and global portions of the network.
    }
    \label{fig:9}
\end{figure}

Finally, we sought to confirm whether ``local'' and ``global'' suppression, which we defined so far by the latency of response, did in fact reflect a spatial property of the network. To do so, we assessed the distance between the center of the target stimulation site and each neuron in all networks and stimulus combinations, and calculated $\R{A}{}$, $\T{A}{}$ and $\D{A}{}$ as a function of this distance. \FIG{9}A-c shows that there were indeed clear spatially defined differences between \dev{}, \msc{} and \std{}. Responses, normalized by the response to stimulation in the recovered networks (\FIG{9}A), clearly spread out farthest in \dev{}, and least in \std{}, despite a nearly indistinguishable response at the immediate stimulation site. For thresholds, $\T[std]{A}{}$ (\FIG{9}B) was much stronger near the stimulation site, and somewhat weaker distally, than either of the less frequent paradigms. A similar pattern emerged with $\D{A}{}$ (\FIG{9}C), though here, the local patch covered a larger area, as we might expect from a synaptic effect.

To quantify this statistically, we split the neurons into local and distal (global) subsets, cutting at a radius of 2 mm from the stimulus site, yielding an average local subset size of 226.8 $\pm$ 13.2 neurons, out of 1000. We then assessed the difference of the mean values of $\T{A}{}$ and $\D{A}{}$ across the two subsets in each sequence, as shown in \FIG{9}D. In keeping with the previously stated analysis, we found that local suppression was higher in \std{} than in \dev{} both for $\T{A}{}$ (T = 482, p = 8.31e-17; one-sided Wilcoxon, n = 120) and for $\D{A}{}$ (T = 735, p = 1.71e-14), indicating that the network response to the non-target stimulus in the \dev{} sequence was indeed kept in check by high local suppression. Similarly, we found that global suppression was greater in \msc{} than in \dev{}, again for both $\T{A}{}$ (T = 1.09e+03, p = 1.37e-11) and $\D{A}{}$ (T = 927, p = 7.27e-13), indicating that the \msc{} sequence produced more suppression across the wider network, leading to smaller target responses than in the \dev{} sequence. Incidentally, this greater global suppression, caused by the varied stimulation paradigm, was also reflected in greater local suppression around the target stimulus site for $\T{A}{}$ (T = 831, p = 1.15e-13) and $\D{A}{}$ (T = 1.46e+03, p = 6.13e-09), providing an additional pathway for a comparatively large response in \dev{}.

@@

\begin{figure}
    \includegraphics[width=\linewidth]{fig-conclusion}
    \caption{%
        Illustration of the origin of DD in our model.
        \textbf{A} Left: In an oddball context, the regular inputs are encoded in the network by high local suppression, which reduces their overall response, leading to low global suppression. Deviant inputs encounter low local and global suppression and therefore evoke a large response. Right: In an irregular context, the network frequently encounters rare stimuli with low local suppression, leading to large overall responses and high global suppression, with these two (global response and suppression) counter-balancing each other. Deviant inputs encounter this high global suppression and therefore evoke smaller responses.
        \textbf{B} Adding additional suppression (e.g., STD) affects these situations differently. Left: In an oddball context, the additional suppression further reduces the regular response, permitting even greater recovery of the wider network. Right: In an irregular context, global suppression is instead increased, and all responses reduced.
    }
    \label{fig:conclusion}
\end{figure}

@@

In this study, we showed that non-synaptic neuronal adaptation can induce true deviance detection in a randomly wired, but spatially organized network of simple leaky integrate-and-fire neurons. Neuronal adaptation achieves this, as illustrated in \FIG{conclusion}A, by suppressing the response of neurons local to the frequently stimulated site. This drastically reduces the response to frequent stimulations at an early stage of signal propagation through the network, allowing the more distant parts of the network to recover from adaptation and respond vigorously to infrequent deviant stimuli. By contrast, a more varied paradigm including inputs at multiple stimulation sites allows local suppression to partly recover between stimulus presentations, resulting in a greater response through greater recruitment of the wider network. As a result, the network becomes more adapted, diminishing the response to each individual stimulus. Deviance detection, then, is a side effect of learning and adapting to the regularity of the frequently presented stimulus, which leaves the wider network less adapted and more responsive than usual.

@@

Our second main finding is that synaptic short-term depression, which acts in similar fashion to neuronal adaptation both in terms of how it is induced and in how it affects activity flow in the network, enhances the DD effect of neuronal adaptation, and vice versa. We demonstrated that this synergy is a result of increased local suppression of the frequently presented stimulus by STD, which decreases global suppression by both mechanisms in oddball sequences and enhances the deviant response, as illustrated in \FIG{conclusion}B. Conversely, without strong local suppression in control sequences, global suppression is increased, leading to smaller individual responses.

@@

\section{Acknowledgments}

This work was supported by World Premier International Research Center Initiative (WPI), MEXT, Japan.

### 2023-03-20 paper-1/manuscript.tex
Change R/T/D from \mathbf to oblique \boldsymbol
Added WPI funding ack
Fixed hat t_j appearance
Aligned fig file naming & fix references in tex

% Common notation
\renewcommand{\R}[3][]{{}^{#1}_{}\mathbf{R}^{#2}_{#3}}
\renewcommand{\T}[3][]{{}^{#1}_{}\mathbf{T}^{#2}_{#3}}
\newcommand{\D}[3][]{{}^{#1}_{}\mathbf{D}^{#2}_{#3}}
\newcommand{\mean}[1]{\overline{#1}}
\newcommand{\Rhat}[3][]{{}^{#1}_{}\mathbf{\widehat{R}}^{#2}_{#3}}

@@

\begin{figure*}%
    \centering
    \includegraphics{fig10.pdf}
    \caption{}
    \label{fig:10}
\end{figure*}
\textbf{Fig. 9} Suppression acts in spatially distinguishable subpopulations.
\textbf{a} Response size $\R{A}{}$ as a function of distance from the center of the stimulation site, evaluated across all networks and stimuli in 200 bins between 0 mm and 6.5 mm, and normalized by the response size in recovered networks (initial trial). Each of the 200 data points represents the mean response across all relevant trials of all neurons in the corresponding distance bin. Left, spatial illustration, using all data with a fictive stimulus location. Right, same data plotted as lines. The vertical line indicates the chosen local/global divider.
\textbf{b} As in panel a, but showing $\T{A}{}$ without normalization.
\textbf{c} As in panel a, but showing $\D{A}{}$ without normalization.
\textbf{d} Mean $\T{A}{}$ in neurons <2 mm from the stimulation site (local) and more distant neurons (global). Data are averaged across relevant neurons and trials in each network and stimulus, leaving 120 data points to be summarized by box plot. Comparisons of interest are between \std{} and \dev{} in the local subset, and between \msc{} and \dev{} in the global subset.
\textbf{e} Corresponding mean $\D{A}{}$ in local and global portions of the network.

Finally, we sought to confirm whether ``local'' and ``global'' suppression, which we defined so far by the latency of response, did in fact reflect a spatial property of the network. To do so, we assessed the distance between the center of the target stimulation site and each neuron in all networks and stimulus combinations, and calculated $\R{A}{}$, $\T{A}{}$ and $\D{A}{}$ as a function of this distance. \FIG{10}a-c shows that there were indeed clear spatially defined differences between \dev{}, \msc{} and \std{}. Responses, normalized by the response to stimulation in the recovered networks (\FIG{10}a), clearly spread out farthest in \dev{}, and least in \std{}, despite a nearly indistinguishable response at the immediate stimulation site. For thresholds, $\T[std]{A}{}$ (\FIG{10}b) was much stronger near the stimulation site, and somewhat weaker distally, than either of the less frequent paradigms. A similar pattern emerged with $\D{A}{}$ (\FIG{10}c), though here, the local patch covered a larger area, as we might expect from a synaptic effect.

To quantify this statistically, we split the neurons into local and distal (global) subsets, cutting at a radius of 2 mm from the stimulus site, yielding an average local subset size of 226.8 $\pm$ 13.2 neurons, out of 1000. We then assessed the difference of the mean values of $\T{A}{}$ and $\D{A}{}$ across the two subsets in each sequence, as shown in \FIG{10}d. In keeping with the previously stated analysis, we found that local suppression was higher in \std{} than in \dev{} both for $\T{A}{}$ (T = 482, p = 8.31e-17; one-sided Wilcoxon, n = 120) and for $\D{A}{}$ (T = 735, p = 1.71e-14), indicating that the network response to the non-target stimulus in the \dev{} sequence was indeed kept in check by high local suppression. Similarly, we found that global suppression was greater in \msc{} than in \dev{}, again for both $\T{A}{}$ (T = 1.09e+03, p = 1.37e-11) and $\D{A}{}$ (T = 927, p = 7.27e-13), indicating that the \msc{} sequence produced more suppression across the wider network, leading to smaller target responses than in the \dev{} sequence. Incidentally, this greater global suppression, caused by the varied stimulation paradigm, was also reflected in greater local suppression around the target stimulus site for $\T{A}{}$ (T = 831, p = 1.15e-13) and $\D{A}{}$ (T = 1.46e+03, p = 6.13e-09), providing an additional pathway for a comparatively large response in \dev{}.

@@

\begin{figure*}%
    \centering
    \includegraphics{fig9.pdf}
    \caption{}
    \label{fig:9}
\end{figure*}
\textbf{Fig. 10} Illustration of the origin of DD in our model.
\textbf{a} Left: In an oddball context, the regular inputs are encoded in the network by high local suppression, which reduces their overall response, leading to low global suppression. Deviant inputs encounter low local and global suppression and therefore evoke a large response. Right: In an irregular context, the network frequently encounters rare stimuli with low local suppression, leading to large overall responses and high global suppression, with these two (global response and suppression) counter-balancing each other. Deviant inputs encounter this high global suppression and therefore evoke smaller responses.
\textbf{b} Adding additional suppression (e.g., STD) affects these situations differently. Left: In an oddball context, the additional suppression further reduces the regular response, permitting even greater recovery of the wider network. Right: In an irregular context, global suppression is instead increased, and all responses reduced.

@@

In this study, we showed that non-synaptic neuronal adaptation can induce true deviance detection in a randomly wired, but spatially organized network of simple leaky integrate-and-fire neurons. Neuronal adaptation achieves this, as illustrated in Figure 10a, by suppressing the response of neurons local to the frequently stimulated site. This drastically reduces the response to frequent stimulations at an early stage of signal propagation through the network, allowing the more distant parts of the network to recover from adaptation and respond vigorously to infrequent deviant stimuli. By contrast, a more varied paradigm including inputs at multiple stimulation sites allows local suppression to partly recover between stimulus presentations, resulting in a greater response through greater recruitment of the wider network. As a result, the network becomes more adapted, diminishing the response to each individual stimulus. Deviance detection, then, is a side effect of learning and adapting to the regularity of the frequently presented stimulus, which leaves the wider network less adapted and more responsive than usual.

@@

Our second main finding is that synaptic short-term depression, which acts in similar fashion to neuronal adaptation both in terms of how it is induced and in how it affects activity flow in the network, enhances the DD effect of neuronal adaptation, and vice versa. We demonstrated that this synergy is a result of increased local suppression of the frequently presented stimulus by STD, which decreases global suppression by both mechanisms in oddball sequences and enhances the deviant response, as illustrated in Figure 10b. Conversely, without strong local suppression in control sequences, global suppression is increased, leading to smaller individual responses.

@@

\bmhead{Funding}
This study was funded by a World Premier International Research Center Initiative research startup grant.

>>>>

% Common notation
\renewcommand{\R}[3][]{{}^{#1}_{}\boldsymbol R^{#2}_{#3}}
\renewcommand{\T}[3][]{{}^{#1}_{}\boldsymbol T^{#2}_{#3}}
\newcommand{\D}[3][]{{}^{#1}_{}\boldsymbol D^{#2}_{#3}}
\newcommand{\mean}[1]{\overline{#1}}
\newcommand{\Rhat}[3][]{{}^{#1}_{}\widehat{\boldsymbol R}^{#2}_{#3}}

@@

\begin{figure*}%
    \centering
    \includegraphics{fig9.pdf}
    \caption{}
    \label{fig:9}
\end{figure*}
\textbf{Fig. 9} Suppression acts in spatially distinguishable subpopulations.
\textbf{a} Response size $\R{A}{}$ as a function of distance from the center of the stimulation site, evaluated across all networks and stimuli in 200 bins between 0 mm and 6.5 mm, and normalized by the response size in recovered networks (initial trial). Each of the 200 data points represents the mean response across all relevant trials of all neurons in the corresponding distance bin. Left, spatial illustration, using all data with a fictive stimulus location. Right, same data plotted as lines. The vertical line indicates the chosen local/global divider.
\textbf{b} As in panel a, but showing $\T{A}{}$ without normalization.
\textbf{c} As in panel a, but showing $\D{A}{}$ without normalization.
\textbf{d} Mean $\T{A}{}$ in neurons <2 mm from the stimulation site (local) and more distant neurons (global). Data are averaged across relevant neurons and trials in each network and stimulus, leaving 120 data points to be summarized by box plot. Comparisons of interest are between \std{} and \dev{} in the local subset, and between \msc{} and \dev{} in the global subset.
\textbf{e} Corresponding mean $\D{A}{}$ in local and global portions of the network.

Finally, we sought to confirm whether ``local'' and ``global'' suppression, which we defined so far by the latency of response, did in fact reflect a spatial property of the network. To do so, we assessed the distance between the center of the target stimulation site and each neuron in all networks and stimulus combinations, and calculated $\R{A}{}$, $\T{A}{}$ and $\D{A}{}$ as a function of this distance. \FIG{9}a-c shows that there were indeed clear spatially defined differences between \dev{}, \msc{} and \std{}. Responses, normalized by the response to stimulation in the recovered networks (\FIG{9}a), clearly spread out farthest in \dev{}, and least in \std{}, despite a nearly indistinguishable response at the immediate stimulation site. For thresholds, $\T[std]{A}{}$ (\FIG{9}b) was much stronger near the stimulation site, and somewhat weaker distally, than either of the less frequent paradigms. A similar pattern emerged with $\D{A}{}$ (\FIG{9}c), though here, the local patch covered a larger area, as we might expect from a synaptic effect.

To quantify this statistically, we split the neurons into local and distal (global) subsets, cutting at a radius of 2 mm from the stimulus site, yielding an average local subset size of 226.8 $\pm$ 13.2 neurons, out of 1000. We then assessed the difference of the mean values of $\T{A}{}$ and $\D{A}{}$ across the two subsets in each sequence, as shown in \FIG{9}d. In keeping with the previously stated analysis, we found that local suppression was higher in \std{} than in \dev{} both for $\T{A}{}$ (T = 482, p = 8.31e-17; one-sided Wilcoxon, n = 120) and for $\D{A}{}$ (T = 735, p = 1.71e-14), indicating that the network response to the non-target stimulus in the \dev{} sequence was indeed kept in check by high local suppression. Similarly, we found that global suppression was greater in \msc{} than in \dev{}, again for both $\T{A}{}$ (T = 1.09e+03, p = 1.37e-11) and $\D{A}{}$ (T = 927, p = 7.27e-13), indicating that the \msc{} sequence produced more suppression across the wider network, leading to smaller target responses than in the \dev{} sequence. Incidentally, this greater global suppression, caused by the varied stimulation paradigm, was also reflected in greater local suppression around the target stimulus site for $\T{A}{}$ (T = 831, p = 1.15e-13) and $\D{A}{}$ (T = 1.46e+03, p = 6.13e-09), providing an additional pathway for a comparatively large response in \dev{}.

@@

\begin{figure*}%
    \centering
    \includegraphics{fig-conclusion.pdf}
    \caption{}
    \label{fig:conclusion}
\end{figure*}
\textbf{Fig. 10} Illustration of the origin of DD in our model.
\textbf{a} Left: In an oddball context, the regular inputs are encoded in the network by high local suppression, which reduces their overall response, leading to low global suppression. Deviant inputs encounter low local and global suppression and therefore evoke a large response. Right: In an irregular context, the network frequently encounters rare stimuli with low local suppression, leading to large overall responses and high global suppression, with these two (global response and suppression) counter-balancing each other. Deviant inputs encounter this high global suppression and therefore evoke smaller responses.
\textbf{b} Adding additional suppression (e.g., STD) affects these situations differently. Left: In an oddball context, the additional suppression further reduces the regular response, permitting even greater recovery of the wider network. Right: In an irregular context, global suppression is instead increased, and all responses reduced.

@@

In this study, we showed that non-synaptic neuronal adaptation can induce true deviance detection in a randomly wired, but spatially organized network of simple leaky integrate-and-fire neurons. Neuronal adaptation achieves this, as illustrated in \FIG{conclusion}a, by suppressing the response of neurons local to the frequently stimulated site. This drastically reduces the response to frequent stimulations at an early stage of signal propagation through the network, allowing the more distant parts of the network to recover from adaptation and respond vigorously to infrequent deviant stimuli. By contrast, a more varied paradigm including inputs at multiple stimulation sites allows local suppression to partly recover between stimulus presentations, resulting in a greater response through greater recruitment of the wider network. As a result, the network becomes more adapted, diminishing the response to each individual stimulus. Deviance detection, then, is a side effect of learning and adapting to the regularity of the frequently presented stimulus, which leaves the wider network less adapted and more responsive than usual.

@@

Our second main finding is that synaptic short-term depression, which acts in similar fashion to neuronal adaptation both in terms of how it is induced and in how it affects activity flow in the network, enhances the DD effect of neuronal adaptation, and vice versa. We demonstrated that this synergy is a result of increased local suppression of the frequently presented stimulus by STD, which decreases global suppression by both mechanisms in oddball sequences and enhances the deviant response, as illustrated in \FIG{conclusion}b. Conversely, without strong local suppression in control sequences, global suppression is increased, leading to smaller individual responses.

@@

\bmhead{Funding}
This work was supported by World Premier International Research Center Initiative (WPI), MEXT, Japan.
